---
title: "RegressionModels_CourseProject"
author: "Joshua"
date: "23 September 2015"
output: pdf_document
---
#Regression Models - Course Project
  
##Executive summary
This report explores the relationship between a set of variables and miles per gallon (MPG) from the mtcars data set and attempts to determine if an automatic or manual transmission better for MPG and quantifies the actual different in MPG. 

Both the simple linear and multiple regression models indicate that manual transmission have on average a higher miles per gallon figure than automatic transmission. In the simple linear regression model between MPG and transmission type alone the increase in miles per gallon is approximately 7 miles per gallon. In the multiple regression model, considering the other predictors weight, horsepower and number of cylinders, the increase is approximately 1.8 MPG when switching from an automatic transmission to a manual one, with all other predictors held constant.

Exploratory data analysis and diagramatic plots are located in the Appendix of this document.

##Data processing
We begin by loading the mtcars dataset and converting some variables to factors.
```{r}
data(mtcars)
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am <- factor(mtcars$am,labels=c("Automatic","Manual"))
colnames(mtcars)[9]<-"TransType"
```
##Analysis
### Simple Linear Regression between mpg and transmission
Is there a different between mpg achieved with manual vs automatic transmissions?
The boxplot (fig 1) in the appendix clearly indicates a higher mean mpg figure when comparing automatic vs manual transmissions, on average about 7 more miles to the gallon. A t-test shows that the difference in means mpg is significant. (The p-value < 0.05).
```{R}
simple_fit<-lm(mpg ~ TransType,mtcars)
simple_fit
t.test(mpg~TransType,mtcars)
```
However we can also see that the adjust R-squared value of the simple linear regression model is 0.3385, meaning a different tranmission only explains 33.85% of the variance in mpg.

### Multiple regression analysis
From the pairs plot (fig 2), several other variables seem to have high correlation with mpg. To find the best model fit build an initial model including all the variables as predictors and perform a stepwise model selection to determine which are the predictors required for the best fit.
The R function step can do this quite easily, it chooses a model based on AIC. (https://en.wikipedia.org/wiki/Akaike_information_criterion)
```{R}
fullmodel <- lm(mpg ~ ., data = mtcars)
bestmodel <- step(fullmodel, direction = "both")
```
From the best model we see that in addition to transmission type, cyl, hp, and wt are all also important variables in predicting mpg. The adjusted R-squared value is .8401 which tells us that 84% of the variability is explained by this model.
Comparing both the simple linear regression model and the best fit model, the p-value is very small, indicating that the two models are significantly different. The combination of weight, horsepower and transmission type explain 84% of the variability in mileage achieved per gallon.
```{R}
anova(simple_fit,bestmodel)
```

### Diagnostics
Looking at a plot of the best model in Figure 3. we can see that:
  1. The Residuals vs Fitted and Scale-Location plots do not have any systematic pattern.
  2. The normal Q-Q plot shows the point generally clustered along the line, indicating the residuals are normally distributed.
  3. There are a few 

Influential points:
Outlier points:
Leverage:

##Appendix
```{R}
library(ggplot2)
```
###Figure 1: Box plot of mpg by transmission type
```{R}
fig1<-ggplot(mtcars, aes(factor(TransType), mpg, fill=factor(TransType))) +
  geom_boxplot() +
  scale_colour_discrete(name = "Type") + 
  scale_fill_discrete(name="Type", breaks=c("0", "1"),labels=c("Automatic", "Manual")) +
  scale_x_discrete(breaks=c("0", "1"), labels=c("Automatic", "Manual")) +
  xlab("Transmission Type: Automatic, Manual")
fig1
```

###Figure 2: Pairs plot
```{r}
fig2<-pairs(mtcars,panel=panel.smooth,col=9,main="Pairs plot of mtcars dataset")
fig2
```

###Figure 3: Plot of best fit model
```{R}
par(mfrow =c(2,2))
plot(bestmodel)
```